How Toronto's homeless death toll over the years shows the dangers of addiction's epidemic?

### 2024-01-24 outputs/paper/paper.qmd
Finished Jan 23 work

---
title: "How Toronto's homeless death toll over the years shows the dangers of addiction's epidemic?"
subtitle: "Data analysis of homeless deaths in Toronto from 2017 to 2023"
author: 
  - Shipeng Zhang
thanks: "Code and data are available at: LINK."
date: 2024-01-22
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

@@

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

@@

file_path <- "../../inputs/data/raw_data.csv"
toronto_homeless_death <- read.csv(file_path)
toronto_homeless_death <- dplyr::select(toronto_homeless_death, -X_id)           
sample_data <- head(toronto_homeless_death, 6)
knitr::kable(sample_data,'markdown',col.names = c("Year of Death", "Cause of Death", "Age Group", "Gender", "Count"),align = 'c',

                      caption = "Sample Table for Causes of Death")
  # Print the table

@@

```{r}
#| label: fig-causes-of-death
#| fig-cap: Causes of death analysis
#| echo: false

ggplot(data = toronto_homeless_death,mapping =  aes(x = Cause_of_death, fill = Age_group)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "lightpink", "lightyellow", "lightgray")) +
  labs(title = "Causes of Death Stacked by Age Group",
       x = "Cause of Death",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but
don't worry about doing that until you have finished every other aspect
of the paper - Quarto will try to make it look nice and the defaults
usually work well once you have enough text.)

#\| label: fig-planes #\| fig-cap: Relationship between wing length and
width #\| echo: false #\| warning: false #\| message: false

analysis_data \<- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data \|\> ggplot(aes(x = width, y = length)) + geom_point(alpha
= 0.8) + theme_minimal() + labs(x = "Wing width (mm)", y = "Wing length
(mm)")

Talk way more about it.

>>>>

---
title: "How Toronto's homeless death toll over the years shows the dangers of addiction's epidemic?"
subtitle: "Data analysis of homeless deaths in Toronto from 2017 to 2023"
author: 
  - Shipeng Zhang
thanks: "Code and data are available at: https://github.com/zero616/Homeless_Death_Analysis."
date: 2024-01-22
date-format: long
abstract: "Homelessness is a well-known problem in Toronto, and this article takes an in-depth look at the tragic reality of homelessness in Toronto. Focuses on the devastating impact of drug addiction by analyzing homeless death toll data from 2017 to 2023. Our top findings reveal a worrying rise in addiction-related homeless deaths over the years. By examining causes of death, age groups and gender, we provide a comprehensive overview revealing the dangers faced by this vulnerable group."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Toronto is currently experiencing an escalating drug poisoning crisis,
with challenges posed by the Covid-19 pandemic further exacerbating the
public health dilemma. This article explores the details behind this
crisis through a comprehensive analysis of homeless deaths from 2017 to
2023.

Drug addiction has a profound impact on the homeless community. Our
result reveal the rapid spread of drug poisoning in Toronto. This
article takes an depth look at the proportion of deaths caused by drug
poisoning and sheds light on the impact of the COVID-19 pandemic on the
drug poisoning crisis.

To examine the impact of the COVID-19 pandemic on the drug poisoning
crisis, this article is divided into the following sections: data,
results, discussion, and conclusion. In the data section, I discuss the
nature of the spreadsheets obtained through the City of Toronto's
OpenDataToronto (Toronto Public Health 2023) and the steps I took to
clean and analyze the data. The Results section displays the trends
discovered during the analysis, and the Discussion section evaluates the
trends and provides insights. Finally, the conclusion summarizes the
main findings of this paper.

@@

The dataset used in this analysis was obtained from the City of
Toronto's OpenDataToronto Library (Gelfand 2022), adhering to ethical
guidelines and ensuring personal anonymity. It includes important
information such as year of death, cause of death, age group, gender and
number of people. To maintain transparency and reproducibility, all
analyzes were performed using the R programming language and visualized
with ggplot2 (Wickham 2016) and tables created with knit (Xie 2023).

Table 1 provides a summary of the dataset outlining the key variables
under consideration: time to death, cause of death, sex and age group.
The time of death started in 2017 and ended in 2023, and the age groups
were divided into \<20, 20-39, 40-59, 60+.

@@

file_path <- "../../inputs/data/raw_data.csv"
toronto_homeless_death <- read.csv(file_path)
toronto_homeless_death <- dplyr::select(toronto_homeless_death, -X_id)   
# Remove rows where 'Gender' is 'Unknown'
toronto_homeless_death <- toronto_homeless_death %>% filter(Gender != 'Unknown')
toronto_homeless_death <- toronto_homeless_death %>% filter(Age_group != 'Unknown')
# Remove rows where 'Cause_of_death' is 'Unknown' or 'Pending'
toronto_homeless_death <- toronto_homeless_death %>%
  filter(Cause_of_death != 'Unknown/Pending')

sample_data <- head(toronto_homeless_death, 6)
knitr::kable(sample_data,'markdown',col.names = c("Year of Death", "Cause of Death", "Age Group", "Gender", "Count"),align = 'c',

                      caption = "Sample Table for Toronto Homeless Death")
  # Print the table

@@

```{r}
total_count_age_data <- toronto_homeless_death %>%
  group_by(Age_group, Cause_of_death) %>%
  summarise(total_count_age = sum(Count), .groups = 'drop')
sample_total_count_by_age_data <- head(total_count_age_data, 6)
knitr::kable(sample_total_count_by_age_data,'markdown',col.names = c("Age Group", "Cause of Death", "Total Count"),align = 'c',

                      caption = "Sample Table for total death number of causes of death by Age Group")
```

```{r}
total_count_data <- toronto_homeless_death %>%
  group_by(Year.of.death, Cause_of_death) %>%
  summarise(total_count = sum(Count), .groups = 'drop')
sample_total_count_by_year_data <- head(total_count_data, 6)
knitr::kable(sample_total_count_by_year_data,'markdown',col.names = c("Year of Death", "Cause of Death", "Total Count"),align = 'c',

                      caption = "Sample Table for total death number of causes of death by Age Group")
```

Figure 1 shows the distribution of causes of death among homeless
people, stratified by age group. This figure provides an impactful
sketch of the leading causes of death among homeless people.

```{r}
#| label: fig-causes-of-death-by-age
#| fig-cap: Causes of death analysis by abe
#| warning: false
#| message: false
#| echo: false

ggplot(data = total_count_age_data,mapping =  aes(x = Cause_of_death,y =total_count_age, fill = Age_group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "lightpink", "lightyellow", "lightgray")) +
  labs(title = "Causes of Death Stacked by Age Group",
       x = "Cause of Death",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig-causes-of-death-by-year
#| fig-cap: Causes of death analysis by year
#| echo: false
#| warning: false
#| message: false
library(ggplot2)
library(dplyr)

# Assuming toronto_homeless_death is your dataset

# Aggregate the data to get the total count for each cause of death in each year

# Create a stacked bar chart
ggplot(data = total_count_data, aes(x = Year.of.death, y = total_count, fill = Cause_of_death)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("lightblue", "lightgreen", "lightpink", "lightyellow", "lightgray","black","orange","purple","yellow","red")) +
  labs(title = "Total Count of Causes of Death Stacked by Year",
       x = "Year of Death",
       y = "Death Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
```

```{r}
#| label: fig-rate-of-death
#| fig-cap: Rate of death Caused by Drug Toxicity Over Time
#| warning: false
#| message: false
#| echo: false
# Filter data for Drug Toxicity
drug_toxicity_data <- total_count_data %>%
  filter(Cause_of_death == "Drug Toxicity")

# Calculate the rate (percentage) of Drug Toxicity deaths for each year
rate_data <- total_count_data %>%
  group_by(Year.of.death) %>%
  summarise(total_deaths = sum(total_count),
            drug_toxicity_deaths = sum(total_count[Cause_of_death == "Drug Toxicity"]),
            rate = (drug_toxicity_deaths / total_deaths) * 100)  # Multiply by 100 to convert to percentage

# Create a line chart for the rate of Drug Toxicity deaths over time
ggplot(data = rate_data, aes(x = Year.of.death, y = rate)) +
  geom_line(color = "red",linewidth = 1.5) +
  geom_text(aes(label = sprintf("%.2f%%", rate)), vjust = -0.5, hjust = 0.5, size = 3) +
  labs(title = "Rate of Deaths Caused by Drug Toxicity Over Time",
       x = "Year of Death",
       y = "Rate (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))
```
